---
title: "Binary Random Forest Classification"
output: html_document
---



```{r setup, include=FALSE}
#Loading Required Libraries
library(randomForest)
library(dplyr)
library(ggplot2)
library(caret)
library(pROC)
library(ROSE)  
library(smotefamily)
#Setting seed for reproducibility 
set.seed(123)
```


```{r Loading and Preparing Data, include=FALSE}
#Loading the crash data
driver_data <- read.csv("data/DriverInvolvement.csv")

#Selecting Crash Severity and Binary Driver Indicators
driver_filtered <- driver_data %>%
  select(Crash_Severity, starts_with("Involving"))

#Converting Yes/No to binary 1/0
driver_filtered[, -1] <- lapply(driver_filtered[, -1], function(x) ifelse(x == "Yes", 1, 0))

#Converting Crash_Severity to binary: 1 = Fatal, 0 = Others
driver_filtered$FatalBinary <- ifelse(driver_filtered$Crash_Severity == "Fatal", 1, 0)

#Converting Target to Factor for Classification
driver_filtered$FatalBinary <- factor(driver_filtered$FatalBinary)

```


```{r Split Into Train/Test, include=FALSE}
#80/20 split
split_index <- createDataPartition(driver_filtered$FatalBinary, p = 0.8, list = FALSE)
train_data <- driver_filtered[split_index, ]
test_data  <- driver_filtered[-split_index, ]
```


```{r Exploratory Analysis – Fatality Rate by Driver Type}
#Preparing long-format data for visual analysis
library(tidyr)

#Converting FatalBinary to numeric for aggregation
driver_filtered$FatalBinary <- as.numeric(as.character(driver_filtered$FatalBinary))

#Selecting only Involving_ columns
involvement_features <- names(driver_filtered)[grepl("Involving_", names(driver_filtered))]

#Creating long-format data
long_df <- driver_filtered %>%
  select(FatalBinary, all_of(involvement_features)) %>%
  pivot_longer(cols = all_of(involvement_features), names_to = "DriverType", values_to = "Involved") %>%
  filter(Involved == 1)

#Summarising fatality rates
fatality_summary <- long_df %>%
  group_by(DriverType) %>%
  summarise(
    FatalCrashes = sum(FatalBinary),
    TotalCrashes = n(),
    FatalityRate = round(FatalCrashes / TotalCrashes, 3)
  ) %>%
  arrange(desc(FatalityRate))

#Plotting for presentation
ggplot(fatality_summary, aes(x = reorder(DriverType, -FatalityRate), y = FatalityRate)) +
  geom_col(fill = "darkred") +
  geom_text(aes(label = FatalityRate), vjust = -0.5, size = 3.5) +
  labs(
    title = "Fatal Crash Rate by Driver Type",
    subtitle = "Driver types with higher proportions of fatal outcomes",
    x = "Driver Type",
    y = "Proportion of Fatal Crashes"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11)
  )
```


```{r Handle Class Imbalance with Downsampling, include=FALSE}
#Downsampling to balance Fatal (1) vs Non-Fatal (0)
train_balanced <- downSample(
  x = train_data[, -which(names(train_data) %in% c("Crash_Severity", "FatalBinary"))],
  y = train_data$FatalBinary
)

#Renaming back to correct column
names(train_balanced)[ncol(train_balanced)] <- "FatalBinary"

#Confirming balanced class sizes
table(train_balanced$FatalBinary)
```


##Model 1: Random Forest with Downsampling

This model is used with the caret package as to downsample the majority class being non-fatal crashes, by creating a balanced training set. This method helps the model gain knowledge by learning in an equal format from both the fatal and non-fatal examples.


```{r train-rf-downsampled}
#Training Random Forest with balanced training set
rf_model <- randomForest(
  FatalBinary ~ .,
  data = train_balanced,
  ntree = 1000,
  importance = TRUE
)

#Printing model summary
print(rf_model)
```


```{r eval-rf-downsampled}
#Probability of class "1" (Fatal)
#rf_probs <- predict(rf_model, newdata = test_data, type = "prob")[, "1"]
rf_preds <- predict(rf_model, newdata = test_data)
confusionMatrix(rf_preds, test_data$FatalBinary, positive = "1")
#Getting predicted probabilities of being fatal
rf_probs <- predict(rf_model, newdata = test_data, type = "prob")[, "1"]
#Plot ROC and calculate AUC
roc_rf <- roc(test_data$FatalBinary, rf_probs)
plot(roc_rf, main = "ROC Curve - Random Forest")
auc(roc_rf)
```


Accuracy: 62.19%
Sensitivity: 67.84%
Specificity: 61.57%
Balanced Accuracy: 64.70%
AUC: 0.6988
Kappa: 0.1226


The Downsampling model is performing well as it detects the minority class (Fatal), by achieving a pretty strong sensitivity of 0.6784, meaning that 68% approximaticaly was correctly identified. The AUC of 0.6988 is an indication of a good overall classifier ability, whereas, the overall accuracy is found to be moderatively good with 62.19% due to the imbalance of classes, which is common when downsampling. Yet, this method is still maintaining a high interpretability and balaced detection rates.


```{r results-by-type-downsampled}
#Adding predicted class (0 or 1) to test set
test_data$Predicted <- predict(rf_model, newdata = test_data)

#Converting factors to numeric
test_data$FatalBinary <- as.numeric(as.character(test_data$FatalBinary))
test_data$Predicted <- as.numeric(as.character(test_data$Predicted))

#Reshaping for per-driver-type comparison
library(tidyr)
driver_columns <- names(test_data)[grepl("Involving_", names(test_data))]

long_test <- test_data %>%
  select(all_of(driver_columns), FatalBinary, Predicted) %>%
  pivot_longer(cols = all_of(driver_columns), names_to = "DriverType", values_to = "Involved") %>%
  filter(Involved == 1)

#Summarizing actual vs predicted
summary_test <- long_test %>%
  group_by(DriverType) %>%
  summarise(
    ActualFatal = sum(FatalBinary),
    PredictedFatal = sum(Predicted),
    Total = n(),
    ActualRate = round(ActualFatal / Total, 3),
    PredictedRate = round(PredictedFatal / Total, 3)
  ) %>%
  arrange(desc(ActualRate))

#Plot
ggplot(summary_test, aes(x = reorder(DriverType, -ActualRate))) +
  geom_col(aes(y = ActualRate), fill = "firebrick", alpha = 0.7, width = 0.6) +
  geom_col(aes(y = PredictedRate), fill = "steelblue", alpha = 0.7, width = 0.3) +
  labs(
    title = "Test Set: Actual vs Predicted Fatality Rates by Driver Type",
    subtitle = "Red = Actual, Blue = Predicted",
    x = "Driver Type Involvement",
    y = "Fatality Rate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The downsampling model has a tendency to over predict rates with all the driver types present, more so, when it comes to males and unlicensed drivers. While it is still capturing general trends, the predictors are of high consistency compared to the actual rates.


```{r importance-downsampled}
#Displaying most influential variables
varImpPlot(rf_model, main = "Variable Importance - Random Forest")
```


The downsampling model shows that "overseas licensed driver" and "female driver" have emerged out to be the most important influencers when ti comes to fatal crash predictions. These variables are contributors to the reducement of uncertainty across decision splits.


##Model 2: Random Forest with Oversampling Using rose package

Using the Rose package, this particular model uses an oversampling method to oversample te minority class "fatal crashes" by creating new synthetic examples. Making it helpful when wanting to improve the model's ability of detecting rara fatal crash cases.


```{r oversampling, include=FALSE}

#Performing oversampling on training data
train_oversampled <- ovun.sample(FatalBinary ~ ., data = train_data, method = "over", N = sum(train_data$FatalBinary == 0) * 2)$data

#Checking class distribution
table(train_oversampled$FatalBinary)
```


```{r train-rf-oversampled}
#Training Random Forest model on oversampled data
rf_model_over <- randomForest(
  FatalBinary ~ .,
  data = train_oversampled[, !names(train_oversampled) %in% "Crash_Severity"],
  ntree = 1000,
  importance = TRUE
)

print(rf_model_over)
```


```{r eval-rf-oversampled}
#Predicting on test set
rf_preds_over <- predict(rf_model_over, newdata = test_data)

#Ensuring factors with same levels
rf_preds_over <- factor(rf_preds_over, levels = c(0, 1))
test_data$FatalBinary <- factor(test_data$FatalBinary, levels = c(0, 1))

#Confusion matrix
confusionMatrix(rf_preds_over, test_data$FatalBinary, positive = "1")
```


```{r eval-rf-oversampled}
#ROC and AUC 
rf_probs_over <- predict(rf_model_over, newdata = test_data, type = "prob")[, "1"]
roc_over <- roc(test_data$FatalBinary, rf_probs_over)
plot(roc_over, main = "ROC Curve - Oversampled Random Forest")
auc(roc_over)
```


Accuracy: 63.64%
Sensitivity: 67.61%
Specificity: 63.21%
Balanced Accuracy: 65.41%
AUC: 0.7O26
Kappa: 0.1321


The ROSE package model has a good balanced classification with some improved metrics over the downsampling method. It has been able to correctly identify about 68% of its fatal crashed, with an AUC of 0.7026, the highest amongst all models, with an a clear indication of strong classifier performance. The moderate accuracy of 63.6% and balanced accuracy of around 65.4% has demonstrated that this model is quite reliable when detecting minority class outcomes while still being able to maintain some sense of stability.


```{r results-by-type-oversampled}
#Adding predicted class to test set
test_data$Predicted_Over <- as.numeric(as.character(rf_preds_over))

#Reshaping data
long_test_over <- test_data %>%
  select(all_of(driver_columns), FatalBinary, Predicted_Over) %>%
  pivot_longer(cols = all_of(driver_columns), names_to = "DriverType", values_to = "Involved") %>%
  filter(Involved == 1)

#Converting to numeric for aggregation
long_test_over$FatalBinary <- as.numeric(as.character(long_test_over$FatalBinary))
long_test_over$Predicted_Over <- as.numeric(as.character(long_test_over$Predicted_Over))

#Summarising actual vs predicted
summary_test_over <- long_test_over %>%
  group_by(DriverType) %>%
  summarise(
    ActualFatal = sum(FatalBinary),
    PredictedFatal = sum(Predicted_Over),
    Total = n(),
    ActualRate = round(ActualFatal / Total, 3),
    PredictedRate = round(PredictedFatal / Total, 3)
  ) %>%
  arrange(desc(ActualRate))

#Plot
ggplot(summary_test_over, aes(x = reorder(DriverType, -ActualRate))) +
  geom_col(aes(y = ActualRate), fill = "firebrick", alpha = 0.7, width = 0.6) +
  geom_col(aes(y = PredictedRate), fill = "steelblue", alpha = 0.7, width = 0.3) +
  labs(
    title = "Test Set: Actual vs Predicted Fatality Rates by Driver Type (Oversampled)",
    subtitle = "Red = Actual, Blue = Predicted",
    x = "Driver Type Involvement",
    y = "Fatality Rate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


This ROSE oversampling model has predicted the fatality rates are a consistant rate for all driver types, more so, when it concerns Males, Unlicensed, Senior drivers, which shows that the sensitivity as improved but the precision is reducing due to the minority class predictions.


```{r importance-oversampled}
#Displaying variable importance for oversampled model
varImpPlot(rf_model_over, main = "Variable Importance - Random Forest (Oversampled)")
```


The ROSE model, the variables like Overseas_licensed_driver, provisional_driver, and femal_driver has been the highest important which is a suggestion that these driver types were the strongest predictors for the identification of fatal crash outcomes.


## Model 3: Random Forest oversampling using SMOTE package

Using SMOTE package is a technique that generates synthethic fatal crash samples, which unlike the ROSE package, creates new cases based on its nearest neightbors, making it ideal to improve the generalisation and reducing the overfitting.


```{r SMOTE Oversampling}
#Loading smotefamily package
library(smotefamily)

#Converting target to numeric (0/1)
train_data$FatalBinary <- as.numeric(as.character(train_data$FatalBinary))

#Removing Crash_Severity
train_smote_input <- train_data[, !names(train_data) %in% "Crash_Severity"]

#Applying SMOTE 
# The last argument (K) is number of neighbors; typically K = 5
smote_result <- SMOTE(train_smote_input[, -which(names(train_smote_input) == "FatalBinary")], 
                      train_smote_input$FatalBinary,
                      K = 5,
                      dup_size = 4)  # 400% over-sampling

#Combining features and new balanced target
train_smote <- smote_result$data
colnames(train_smote)[ncol(train_smote)] <- "FatalBinary"

#Converting FatalBinary back to factor
train_smote$FatalBinary <- factor(train_smote$FatalBinary)

#Training Random Forest
rf_model_smote <- randomForest(
  FatalBinary ~ .,
  data = train_smote,
  ntree = 1000,
  importance = TRUE
)

#Predicting on test set
rf_preds_smote <- predict(rf_model_smote, newdata = test_data)
rf_preds_smote <- factor(rf_preds_smote, levels = c(0, 1))
test_data$FatalBinary <- factor(test_data$FatalBinary, levels = c(0, 1))

#Confusion matrix
confusionMatrix(rf_preds_smote, test_data$FatalBinary, positive = "1")

#AUC and ROC
rf_probs_smote <- predict(rf_model_smote, newdata = test_data, type = "prob")[, "1"]
roc_smote <- roc(test_data$FatalBinary, rf_probs_smote)
plot(roc_smote, main = "ROC Curve - SMOTE Random Forest")
auc(roc_smote)
```


Accuracy: 84.14%
Sensitivity: 20.66%
Specificity: 91.18%
Balanced Accuracy: 55.92%
AUC: 0.6838
Kappa: 0.1182

The oversampling using SMOTE package model has shown a very high accuracy of 84.1%, which is quite misleading since there is a class imbalance, thus meaning its performance for detecting fatal crashes is quite poor, while having at the same time a low sensitivity of 20.66%. This is a clear statement that this model misses around 4 on 5 fatal crashes. The AUC stand around 0.6838 which is found to also be the lowest amongst all models, hence a weak discriminatory power. Despite the strong results for the majority class, this model is in our minds not a reliable one when it comes to the identification of fatal outcomes, which is here our main analysis goal.


```{r results-by-type smote}
#Adding predicted class to test set
test_data$Predicted_SMOTE <- as.numeric(as.character(rf_preds_smote))

#Ensuring FatalBinary is numeric (if not already)
test_data$FatalBinary <- as.numeric(as.character(test_data$FatalBinary))

#Reshaping test data for per-driver-type analysis
library(tidyr)
driver_columns <- names(test_data)[grepl("Involving_", names(test_data))]

long_test_smote <- test_data %>%
  select(all_of(driver_columns), FatalBinary, Predicted_SMOTE) %>%
  pivot_longer(cols = all_of(driver_columns),
               names_to = "DriverType",
               values_to = "Involved") %>%
  filter(Involved == 1)

#Summarising actual vs predicted
summary_test_smote <- long_test_smote %>%
  group_by(DriverType) %>%
  summarise(
    ActualFatal = sum(FatalBinary),
    PredictedFatal = sum(Predicted_SMOTE),
    Total = n(),
    ActualRate = round(ActualFatal / Total, 3),
    PredictedRate = round(PredictedFatal / Total, 3)
  ) %>%
  arrange(desc(ActualRate))

#Plot comparison
ggplot(summary_test_smote, aes(x = reorder(DriverType, -ActualRate))) +
  geom_col(aes(y = ActualRate), fill = "firebrick", alpha = 0.7, width = 0.6) +
  geom_col(aes(y = PredictedRate), fill = "steelblue", alpha = 0.7, width = 0.3) +
  labs(
    title = "Test Set: Actual vs Predicted Fatality Rates by Driver Type (SMOTE)",
    subtitle = "Red = Actual, Blue = Predicted",
    x = "Driver Type Involvement",
    y = "Fatality Rate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


This SMOTE model has a close alignment with the actual fatality rates for all driver types, including the Male and Young drivers, which is an indication of an improvement regarding balance but with a bit of underprediction for some categories.


```{r importance-smote}
varImpPlot(rf_model_smote, main = "Variable Importance - Random Forest (SMOTE)")
```


The SMOTE model shows that "Overseas Licensed Driver" and "Female Driver" have been identified as the most influent predictorrs in the determination of fatal crash outcomes.


## Model 4: Random Forest with Upsampling (caret)


This model uses basic upsampling techniques via caret package to help balance the dataset while duplicating fatal crash cases. This method helps improve the sensitivity by making sure the model is able to see more fatal examples, therefore, detecting rare but cricital cases better.

```{r upsampling-prep}
#Ensuring FatalBinary is a factor
train_data$FatalBinary <- factor(train_data$FatalBinary)

#Displaying original training distribution
cat("Training set dimensions:", dim(train_data), "\n")
cat("Test set dimensions:", dim(test_data), "\n")
cat("Training set class distribution before upsampling:\n")
print(table(train_data$FatalBinary))

#Upsampling the minority class (Fatal = 1)
train_balanced_up <- upSample(
  x = train_data[, names(train_data) != "FatalBinary"],
  y = train_data$FatalBinary,
  yname = "FatalBinary"
)

#Displaying upsampled class distribution
cat("Training set dimensions after upsampling:", dim(train_balanced_up), "\n")
cat("Training set class distribution after upsampling:\n")
print(table(train_balanced_up$FatalBinary))
```


```{r upsampling-model}
#Removing Crash_Severity from upsampled training data
train_balanced_up <- train_balanced_up[, !names(train_balanced_up) %in% "Crash_Severity"]

#Training Random Forest on upsampled data
rf_model_upsample <- randomForest(
  FatalBinary ~ .,
  data = train_balanced_up,
  ntree = 1000,
  importance = TRUE
)

#Removing Crash_Severity from test_data if it exists
if ("Crash_Severity" %in% colnames(test_data)) {
  test_data <- test_data %>% select(-Crash_Severity)
}

#Predicting on test set
rf_preds_upsample <- predict(rf_model_upsample, newdata = test_data)
rf_preds_upsample <- factor(rf_preds_upsample, levels = c(0, 1))
test_data$FatalBinary <- factor(test_data$FatalBinary, levels = c(0, 1))
```


```{r upsampling-eval}
#Confusion Matrix
rf_preds_upsample <- factor(rf_preds_upsample, levels = c("0", "1"))
test_data$FatalBinary <- factor(test_data$FatalBinary, levels = c("0", "1"))
cm_up <- confusionMatrix(rf_preds_upsample, test_data$FatalBinary, positive = "1")
print(cm_up)

#ROC and AUC
# ROC and AUC
rf_probs_upsample <- predict(rf_model_upsample, newdata = test_data, type = "prob")[, "1"]
roc_upsample <- roc(test_data$FatalBinary, rf_probs_upsample)

#Proper ROC plot with AUC and threshold
plot(
  roc_upsample,
  main = "ROC Curve - Upsampled Random Forest",
  print.auc = TRUE,        #Display AUC on plot
  auc.polygon = TRUE,      #Shade under the curve
  grid = TRUE,             #Add background grid
  col = "#FF914D",         #Curve color
  lwd = 2,                 #Line width
  print.thres = TRUE       #Show thresholds on the curve
)

# Print AUC in console
print(auc(roc_upsample))
```


Accuracy: 62.19%
Sensitivity: 67.84%
Specificity: 61.57%
Balanced Accuracy: 64.70%
AUC: 0.7014
Kappa: 0.1226

This model using the upsampling method with caret package has delivered the highest sensitivity of 67.84%, meaning that its the best when it comes to identify fatal crashes among all the models, which is a clear direct link to our analysis goal. Even though, the overall accuracy is 62.19%, the lowest out there, this is an expected outcomes as there is a class imbalance. The AUC of 0.7014 is also one confirmation that the model is able to distinguish between classes. Despite, the moderate Kappa, the model has achieve a practical balance when it comes to detecting fatal and non fatal cases, showcasing its strong use in real world scenarios.


```{r upsampling-plot}
# Add predicted class to test set
test_data$Predicted_Up <- as.numeric(as.character(rf_preds_upsample))

# Convert FatalBinary to numeric for plotting
test_data$FatalBinary <- as.numeric(as.character(test_data$FatalBinary))

# Reshape test data to long format
long_test_up <- test_data %>%
  select(all_of(driver_columns), FatalBinary, Predicted_Up) %>%
  pivot_longer(cols = all_of(driver_columns), names_to = "DriverType", values_to = "Involved") %>%
  filter(Involved == 1)

# Summarize actual vs predicted fatality rates
summary_test_up <- long_test_up %>%
  group_by(DriverType) %>%
  summarise(
    ActualFatal = sum(FatalBinary),
    PredictedFatal = sum(Predicted_Up),
    Total = n(),
    ActualRate = round(ActualFatal / Total, 3),
    PredictedRate = round(PredictedFatal / Total, 3)
  ) %>%
  arrange(desc(ActualRate))

# Plot
ggplot(summary_test_up, aes(x = reorder(DriverType, -ActualRate))) +
  geom_col(aes(y = ActualRate), fill = "firebrick", alpha = 0.7, width = 0.6) +
  geom_col(aes(y = PredictedRate), fill = "steelblue", alpha = 0.7, width = 0.3) +
  labs(
    title = "Test Set: Actual vs Predicted Fatality Rates by Driver Type (Upsampled)",
    subtitle = "Red = Actual, Blue = Predicted",
    x = "Driver Type Involvement",
    y = "Fatality Rate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


The upsampling model has a tendency to overpredict the number of fatalities for all driver types, more so, regarding the young and unlicensed drivers, which shows us that there is an improvement for the sensivity for a trade-off on the overestimation.



```{r Importance Upsampled}
#Plotting variable importance for the Upsampled Random Forest model
varImpPlot(rf_model_upsample, main = "Variable Importance - Random Forest (Upsampled)")
```


```{r Bar chart Importance Upsampling}
#Extracting variable importance
var_imp <- importance(rf_model_upsample)
var_imp_df <- data.frame(Variable = rownames(var_imp), 
                         MeanDecreaseGini = var_imp[, "MeanDecreaseGini"])

#Sorting by importance
library(dplyr)
var_imp_df <- var_imp_df %>% arrange(desc(MeanDecreaseGini))

#Ploting bar chart
library(ggplot2)
ggplot(var_imp_df, aes(x = reorder(Variable, MeanDecreaseGini), y = MeanDecreaseGini)) +
  geom_col(fill = "#FF7F0E") +  
  coord_flip() +
  labs(
    title = "Top Predictors – Random Forest (Upsampled)",
    x = "Driver Type",
    y = "Mean Decrease in Gini"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )
```


The upsampling model has demonstrated that driver types like overseas licensed, provisional, and female drivers are the most influent on the model's predictions. This is important as it highlights key groups that are considered significant contributors to the fatal crash outcomes.


##Conclusion
This analysis enables the exploration related to impact of class imbalance when predicting fatal crash outcomes using Random Forest classifiers. Three distinct models were utilised: One being trained on a downsampled dataset using the caret package, another on an oversampled dataset using the rose package, a third using synthetic sampling with the smote package, and a fourth using the basic upsampling with caret package.

All the models have showcased an improved detection concerning the minority class (fatal crashes), yet some trade-offs between the accuracy, sensitivity, and AUC did occur. The oversampling model was able to achieve the highest accuracy of around 84.1% but had the lowest sensitivity of 0.275, meaning that its ability to identify fatal crashes correctly was sometimes false. The Rose model was able to get the strongest AUC of 0.7026, this indicates a better performance of classification overall, while maintaining a good sensitivity of 0.676. The downsampling model had a balanced senstivity of 0.603 and a pretty descent accuracy of 65.6%, but sadly was not able to outperform the others in any key important metric.

The upsampling model, on the other hand, had got the highest sensitivity of all, around 0.678, with a good competitive AUC of 0.7018. These values show us that a strong performance in detecting fatal crashes was present, one of our primary goals during this analysis. While its accuracy was found to be slightly lower 62.2%, this method retained all fatal crashes while avoiding the discarity of non-fatal examples, thus, providing a more practical and interpretable model that had a balanced performance in all.

Feature analysis was able to highlight similar key driver types that influenced the fatal outcomes, meaning that the patterns and trends found in each model were well captured.

In conclusion, even though each of the models had their own particular strengths, the upsampling model proved to be the most appropriate balance between the important key metrics being sensitivity, AUC, and interpretability, hence, making it the obvious choice for this scenario. The aim of detecting fatal crash risks in an effective manner, while supporting policy decisions based on driver-related risk factors is valid through this model.
