---
title: "Logistic Regression Baseline"
output: html_document
---

```{r}
#install.packages('leaps')
library(leaps)
library(dplyr)
library(ggplot2)
library(glmnet)
library(caret)
library(pROC)
library(car)
library(readr)
```

```{r}
# Data preparation
# Step 1: Load required packages
library(caret)    # For upsampling and data splitting
library(dplyr)    # For data manipulation

# Step 2: Set seed for reproducibility
set.seed(123)

# Step 3: Load the dataset
data <- read.csv("data/DriverInvolvement.csv")
cat("Initial dataset dimensions:", dim(data), "\n")

# Step 4: Select relevant columns (Crash_Severity and Involving_* indicators)
# Retain Crash_Severity and columns starting with "Involving_"
data <- data %>%
  select(Crash_Severity, starts_with("Involving_"))
cat("Selected columns:", names(data), "\n")

# Step 5: Convert Yes/No to 1/0 for Involving_* columns
data[, -1] <- lapply(data[, -1], function(x) ifelse(x == "Yes", 1, 0))
cat("Sample of binarized data:\n")
print(head(data[, 1:3]))  # Show first few columns for verification

# Step 6: Convert Crash_Severity to binary FatalBinary (1 = Fatal, 0 = Others)
data$FatalBinary <- factor(ifelse(data$Crash_Severity == "Fatal", 1, 0))
cat("Initial class distribution (FatalBinary):\n")
print(table(data$FatalBinary))

# Step 7: Remove original Crash_Severity column
data <- data %>% select(-Crash_Severity)

# Step 8: Split into train (80%) and test (20%) sets
split_index <- createDataPartition(data$FatalBinary, p = 0.8, list = FALSE)
train <- data[split_index, ]
test <- data[-split_index, ]
cat("Training set dimensions:", dim(train), "\n")
cat("Test set dimensions:", dim(test), "\n")
cat("Training set class distribution before upsampling:\n")
print(table(train$FatalBinary))

# Step 9: Upsample the minority class (Fatal, 1) in the training set
train_balanced <- upSample(
  x = train[, names(train) != "FatalBinary"],
  y = train$FatalBinary,
  yname = "FatalBinary"
)
cat("Training set dimensions after upsampling:", dim(train_balanced), "\n")
cat("Training set class distribution after upsampling:\n")
print(table(train_balanced$FatalBinary))
str(train_balanced)
```

```{r}
# create a training dataset
cleaned_data.tr <-  select(train_balanced,
                           Involving_Male_Driver,
                           Involving_Female_Driver,
                           Involving_Young_Driver_16.24,
                           Involving_Senior_Driver_60plus,
                           Involving_Provisional_Driver,
                           Involving_Overseas_Licensed_Driver,
                           Involving_Unlicensed_Driver,
                           FatalBinary
                           )

# factorize the cleaned data
transformed_data.tr <- transform(cleaned_data.tr,
                                 Involving_Male_Driver = factor(Involving_Male_Driver),
                                 Involving_Female_Driver =factor(Involving_Female_Driver),
                                 Involving_Young_Driver_16.24 = factor(Involving_Young_Driver_16.24),
                                 Involving_Senior_Driver_60plus = factor(Involving_Senior_Driver_60plus),
                                 Involving_Provisional_Driver = factor(Involving_Provisional_Driver),
                                 Involving_Overseas_Licensed_Driver = factor(Involving_Overseas_Licensed_Driver),
                                 Involving_Unlicensed_Driver = factor(Involving_Unlicensed_Driver),
                                 FatalBinary = factor(FatalBinary)
                                 )

# create a test dataset
cleaned_data.ts <-  select(test, Involving_Male_Driver,
                           Involving_Female_Driver,
                           Involving_Young_Driver_16.24,
                           Involving_Senior_Driver_60plus,
                           Involving_Provisional_Driver,
                           Involving_Overseas_Licensed_Driver,
                           Involving_Unlicensed_Driver,
                           FatalBinary
                           )

# factorize the cleaned dataset
transformed_data.ts <- transform(cleaned_data.ts,
                Involving_Male_Driver = factor(Involving_Male_Driver),
                Involving_Female_Driver = factor(Involving_Female_Driver),
                Involving_Young_Driver_16.24 = factor(Involving_Young_Driver_16.24),
                Involving_Senior_Driver_60plus = factor(Involving_Senior_Driver_60plus),
                Involving_Provisional_Driver = factor(Involving_Provisional_Driver),
                Involving_Overseas_Licensed_Driver = factor(Involving_Overseas_Licensed_Driver),
                Involving_Unlicensed_Driver = factor(Involving_Unlicensed_Driver),
                FatalBinary = factor(FatalBinary)
                )

```

```{r}
# check datasets
# training dataset
str(transformed_data.tr)
summary(transformed_data.tr)
colSums(is.na(transformed_data.tr))

# test dataset
str(transformed_data.ts)
summary(transformed_data.ts)
colSums(is.na(transformed_data.ts))
```
```{r}
# commented out as this part is not necessary

#transformed_data$sev <- ifelse(transformed_data$Crash_Severity == #'Fatal', 
#                1, # For Fatal Crash
#                0 # For Other Crash
#                )
# investigate transformed_data
# str(transformed_data)

```


```{r}
# model creation
mod.1 <- regsubsets(FatalBinary ~ .,
                    data = transformed_data.tr[,],
                    method = 'backward')

mod.1.sum <- summary(mod.1)
mod.1.sum # print result

```

```{r}
# print up to third variables selected
selected_vars <- names(coef(mod.1, id = 3))[-1]
selected_vars

```

```{r}
# model creation based on the selected variables
mod.gl <- glm(FatalBinary ~ Involving_Female_Driver
              + Involving_Provisional_Driver
              + Involving_Overseas_Licensed_Driver,
              data = transformed_data.tr,
              family = 'binomial')

# prediction
pr.gl <- predict(mod.gl,
                 newdata = transformed_data.ts,
                 type = 'response') # Output is probability

summary(pr.gl) # print out the summary

```

```{r}
# Draw a ROC curve
roc_obj <- roc(transformed_data.ts$FatalBinary, pr.gl)
plot(roc_obj, 
     main = "Backward Propagation ROC Curve",
     print.auc = TRUE,        # Display AUC value on the plot
     auc.polygon = TRUE,      # Shade the area under the curve
     grid = TRUE,             # Add a grid for readability
     col = "#FF914D",         # Color of the ROC curve
     lwd = 2,                 # Line width of the ROC curve
     print.thres = TRUE)       # Display threshold values on the curve)
auc(roc_obj)

coords(roc_obj, x = "best", best.method = "youden", transpose = FALSE)

sev.pr <- ifelse(pr.gl > 0.5747412, # Threshold of 0.1, 0.07591005
                1, # This classified as fatal
                0) # This classified as others

# Confusion Matrix - truth vs prediction
table(transformed_data.ts$FatalBinary, # Truth
      sev.pr) # Prediction
```


```{r}
# calculate accuracy
accuracy_3var <- (2992+171)/(2992+851+255+171)
accuracy_3var
```

```{r}
# 1. Fit the full logistic model
full_glm <- glm(FatalBinary ~ Involving_Male_Driver
                + Involving_Female_Driver
                + Involving_Young_Driver_16.24
                + Involving_Senior_Driver_60plus
                + Involving_Provisional_Driver
                + Involving_Overseas_Licensed_Driver
                + Involving_Unlicensed_Driver,
                data = transformed_data.tr, family = binomial)

# 2. Perform backward stepwise selection based on AIC
step_glm <- step(full_glm, direction = "backward", trace = FALSE)
summary(step_glm)

# prediction
pr.full_glm <- predict(full_glm,
                 newdata = transformed_data.ts,
                 type = 'response') # Output is probability

summary(pr.full_glm) # print out the summary

```

plot(roc_obj,
     main = "ROC Curve for KNN Model (Fatal vs. Non-Fatal)",
     print.auc = TRUE,        # Display AUC value on the plot
     auc.polygon = TRUE,      # Shade the area under the curve
     grid = TRUE,             # Add a grid for readability
     col = "#FF914D",         # Color of the ROC curve
     lwd = 2,                 # Line width of the ROC curve
     print.thres = TRUE       # Display threshold values on the curve
)
roc_obj$auc

```{r}
# Draw a ROC curve
roc_obj_2 <- roc(transformed_data.ts$FatalBinary, pr.full_glm)
plot(roc_obj_2, 
     main = "Fully Fitted ROC Curve",
    print.auc = TRUE,        # Display AUC value on the plot
     auc.polygon = TRUE,      # Shade the area under the curve
     grid = TRUE,             # Add a grid for readability
     col = "#FF914D",         # Color of the ROC curve
     lwd = 2,                 # Line width of the ROC curve
     print.thres = TRUE       # Display threshold values on the curve
     )
auc(roc_obj_2)

coords(roc_obj_2, x = "best", best.method = "youden", transpose = FALSE)

sev.pr_2 <- ifelse(pr.full_glm > 0.59351, # Threshold of 0.1, 0.07591005
                1, # This classified as fatal
                0) # This classified as others

# Confusion Matrix - truth vs prediction
table(transformed_data.ts$FatalBinary, # Truth
      sev.pr_2) # Prediction
```

```{r}
# calculate accuracy
accuracy_full <- (3053+162)/(3053+790+264+162)
accuracy_full
```


```{r}
#install.packages("car")
#install.packages("readr")
vif_values <- vif(mod.gl)
print(vif_values)
print("**********************************")
vif_values_glmFull <- vif(full_glm)
print(vif_values_glmFull)

```
```{r}
pairs(data)
```

```{r}
# Function to calculate balanced accuracy
calculate_balanced_accuracy <- function(conf_matrix) {
  recall_0 <- conf_matrix[1, 1] / (conf_matrix[1, 1] + conf_matrix[1, 2]) # Recall for Class 0
  recall_1 <- conf_matrix[2, 2] / (conf_matrix[2, 2] + conf_matrix[2, 1]) # Recall for Class 1
  return((recall_0 + recall_1) / 2)
}

conf_matrix_backward <- matrix(c(2992, 851, 255, 171), nrow = 2, byrow = TRUE) # mod.gl
conf_matrix_full <- matrix(c(3053, 790, 264, 162), nrow = 2, byrow = TRUE) # full_glm

# Calculate balanced accuracy
balanced_acc_backward <- calculate_balanced_accuracy(conf_matrix_backward)
balanced_acc_full <- calculate_balanced_accuracy(conf_matrix_full)

# Print results
cat("Balanced Accuracy (mod.gl):", round(balanced_acc_backward, 4), "\n")
cat("Balanced Accuracy (full_glm):", round(balanced_acc_full, 4), "\n")
```
```{r}
# Function to calculate sensitivity (recall) for each class
calculate_sensitivity <- function(conf_matrix) {
  # Sensitivity for Class 0 (Non-Fatal)
  sens_0 <- conf_matrix[1, 1] / (conf_matrix[1, 1] + conf_matrix[1, 2])
  # Sensitivity for Class 1 (Fatal)
  sens_1 <- conf_matrix[2, 2] / (conf_matrix[2, 2] + conf_matrix[2, 1])
  return(list(Sensitivity_NonFatal = sens_0, Sensitivity_Fatal = sens_1))
}

conf_matrix_backward <- matrix(c(2992, 851, 255, 171), nrow = 2, byrow = TRUE) # mod.gl
conf_matrix_full <- matrix(c(3053, 790, 264, 162), nrow = 2, byrow = TRUE) # full_glm

# Calculate sensitivity for both models
sens_backward <- calculate_sensitivity(conf_matrix_backward)
sens_full <- calculate_sensitivity(conf_matrix_full)

# Print results
cat("Backward Propagation Model (mod.gl):\n")
cat("Sensitivity (Non-Fatal, Class 0):", round(sens_backward$Sensitivity_NonFatal, 4), "\n")
cat("Sensitivity (Fatal, Class 1):", round(sens_backward$Sensitivity_Fatal, 4), "\n\n")
cat("Fully Fitted Model (full_glm):\n")
cat("Sensitivity (Non-Fatal, Class 0):", round(sens_full$Sensitivity_NonFatal, 4), "\n")
cat("Sensitivity (Fatal, Class 1):", round(sens_full$Sensitivity_Fatal, 4), "\n")
```
```{r}
# Function to calculate Kappa from a confusion matrix
calculate_kappa <- function(conf_matrix, actual_labels) {
  # Overall Accuracy (Po)
  total <- sum(conf_matrix)
  po <- sum(diag(conf_matrix)) / total
  
  # Expected Agreement (Pe)
  class_counts <- table(actual_labels)
  actual_props <- class_counts / sum(class_counts)
  predicted_counts <- colSums(conf_matrix)
  predicted_props <- predicted_counts / sum(predicted_counts)
  pe <- sum(actual_props * predicted_props)
  
  # Kappa = (Po - Pe) / (1 - Pe)
  kappa <- (po - pe) / (1 - pe)
  
  return(kappa)
}

# Confusion matrix for Backward Propagation Model (mod.gl)
conf_matrix_backward <- matrix(c(2992, 851, 255, 171), nrow = 2, byrow = TRUE)
dimnames(conf_matrix_backward) <- list("Actual" = c("0", "1"), "Predicted" = c("0", "1"))

# Confusion matrix for Fully Fitted Model (full_glm)
conf_matrix_full <- matrix(c(3053, 790, 264, 162), nrow = 2, byrow = TRUE)
dimnames(conf_matrix_full) <- list("Actual" = c("0", "1"), "Predicted" = c("0", "1"))

# Calculate Kappa for both models
kappa_backward <- calculate_kappa(conf_matrix_backward, transformed_data.ts$FatalBinary)
kappa_full <- calculate_kappa(conf_matrix_full, transformed_data.ts$FatalBinary)

# Print results
cat("Kappa for Backward Propagation Model (mod.gl):", round(kappa_backward, 4), "\n")
cat("Kappa for Fully Fitted Model (full_glm):", round(kappa_full, 4), "\n")
```
The result of each model is like below.
[Backward Propagation Model]
**Accuracy**: 0.7409 
**Sensitivity(Non-Fatal, Class 0)**: 0.7786
**Sensitivity(Fatal, Class 1)**: 0.4014
**AUC**: 0.626
**Balanced Accuracy**: 0.59 

[Fully Fitted Model]
**Accuracy**: 0.7531
**Sensitivity(Non-Fatal, Class 0)**: 0.7944
**Sensitivity(Fatal, Class 1)**: 0.3803
**AUC**: 0.679
**Balanced Accuracy**: 0.5874

According to the result above, our model succeeded in predicting Non-Fatal class very well(approximately 80 %), while Fatal class sensitivity is low (approximately 40 %), meaning it is not good for predicting Fatal class. Although the accuracy demonstrated around 75%, this might be attributed to the unbalanced dataset, as balanced accuracy remained low (approximately 0.6). In terms of AUC, it was improved by 0.53, using all of the variables