---
title: "Multiclass Random Forest Baseline"
output: html_document
---
# Overview
This notebook builds a Random Forest model to predict crash severity using driver involvement data. The dataset is imbalanced, and we use upsampling to balance the training data. We evaluate the model with a confusion matrix, feature importance, and ROC curves.

```{r}
library(randomForest)
library(caret)
library(dplyr)
library(smotefamily)
library(pROC)
library(ggplot2)
library(corrplot)
```
Load the dataset, remove irrelevant columns, and prepare variables for modeling.

- Read DriverInvolvement.csv into a data frame.
- Remove columns like Crash_Police_Region, Crash_Year, and casualty counts.
- Check data structure and summary.
- Confirm no missing values.
- Categorical variables to factors, then encode "Involving_" columns as numeric (Yes = 1, No = 0).
```{r}
data= read.csv("data/DriverInvolvement.csv")

# removing irrelevant columns
data = subset(data, select=-c(Crash_Police_Region, Crash_Year, Count_Crashes,	Count_Casualty_Fatality,	Count_Casualty_Hospitalised,	Count_Casualty_MedicallyTreated,	Count_Casualty_MinorInjury,	Count_Casualty_All))
str(data)
summary(data)
anyNA(data)
```
```{r}
# Convert categorical variables to factors
data = data %>%
  mutate(across(c(Crash_Severity, Involving_Male_Driver,
                  Involving_Female_Driver, Involving_Young_Driver_16.24,
                  Involving_Senior_Driver_60plus, Involving_Provisional_Driver,
                  Involving_Overseas_Licensed_Driver, Involving_Unlicensed_Driver), as.factor))

# Convert Involving columns to numeric random forest ingestion
data <- data %>%
  mutate(across(starts_with("Involving_"), ~ as.numeric(. == "Yes")))
```

Split the data into training (80%) and testing (20%) sets for model training and evaluation.

- Set seed for reproducibility.
- Use stratified sampling to maintain class distribution in Crash_Severity.
- Create training and testing datasets.
```{r}
set.seed(123)  # For reproducibility

# Create training and testing split
trainIndex = createDataPartition(data$Crash_Severity, p = 0.8, list = FALSE)
trainData = data[trainIndex, ]
testData = data[-trainIndex, ]
```
There is a clear unbalance in the classes whereby Fatal only has 1708 instances
```{r}
table(trainData$Crash_Severity)
```

To balance out the dataset, we have options between:
1. Oversampling
  a. Synthetic generation
  b. Simple upsampling by randomly duplicating entries
2. Downsampling

Simple upsampling is applied only to the training dataset because:
1. It maintains the pattern and characteristics of the dataset instead of potentially removing important information/patterns
2. Simple upsampling will not introduce bias by arbitrarily adding new information which is not correct
3. Upsampling only the training dataset ensures that the testing dataset accurately reflects real-life scenario.
```{r}
train_balanced <- upSample(
  x = trainData %>% select(-Crash_Severity),  # Predictors
  y = trainData$Crash_Severity,               # Target variable
  yname = "Crash_Severity"                    # Name of the target column
)

# Confirm balanced class sizes
table(train_balanced$Crash_Severity)
```

```{r}
# Correlation Heatmap
corr <- cor(train_balanced[, grepl("Involving_", names(train_balanced))])
corrplot(corr, method = "color", title = "Correlation of Driver Involvement Features")
```


Compare class distributions before and after upsampling.

- Plot bar chart for original training data (blue).
- Plot bar chart for balanced training data (green).
```{r}
# Plot original class distribution
ggplot(trainData, aes(x = Crash_Severity)) +
  geom_bar(fill = "#ff7300") +
  labs(title = "Class Distribution") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Plot balanced class distribution
ggplot(train_balanced, aes(x = Crash_Severity)) +
  geom_bar(fill = "green") +
  labs(title = "Balanced Class Distribution") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
Training random forest model on balanced training data
```{r}
rf_model = randomForest(Crash_Severity ~ . , data = train_balanced,
                         ntree = 1000,  # Number of trees
                         mtry = sqrt(ncol(train_balanced) - 1),  # Default for classification
                         importance = TRUE,
                         )
```

Examine model performance on training data and feature importance.

- Print model summary (OOB error, confusion matrix).
- Display feature importance table.
- Plot variable importance to identify key predictors.
```{r}
# Print model summary
print(rf_model)

# Displaaeture importance
importance(rf_model)

# Plot variabble imporance
varImpPlot(rf_model)
```
Assess model performance on the test set using a confusion matrix.

- Predict Crash_Severity for test data.
- Compute confusion matrix and performance metrics (accuracy, sensitivity, etc.).
```{r}
predictions = predict(rf_model, newdata = testData)

# Compuute confusion matrix and statistics
confusionMatrix(predictions, testData$Crash_Severity)
```

Compare actual and predicted class counts using a grouped bar chart.

- Extract confusion matrix.
- Calculate actual and predicted counts.
- Create a bar chart comparing actual vs. predicted outcomes.
- Save the chart as a PNG file.
```{r}
# Extract confusion matrix
conf_matrix <- confusionMatrix(predictions, testData$Crash_Severity)$table

# Calculate actual and predicted counts
actual_counts <- table(testData$Crash_Severity)
predicted_counts <- table(predictions)

classes <- levels(testData$Crash_Severity)

# Create data frame for bar chart
results_data <- data.frame(
  Outcome = rep(classes, 2),
  Count = c(actual_counts, predicted_counts),
  Type = rep(c("Actual", "Predicted"), each = length(classes))
)

# Plot grouped bar chart
ggplot(results_data, aes(x = Outcome, y = Count, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Actual vs. Predicted Outcomes (Random Forest)",
    x = "Outcome",
    y = "Count"
  ) +
  scale_fill_manual(values = c("Actual" = "#4ECDC4", "Predicted" = "#FF6B6B")) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.title = element_blank(),
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# Save the bar chart
ggsave("actual_vs_predicted_bar.png", width = 8, height = 6, dpi = 300)
```

Evaluate model performance using multiclass ROC curves (one-vs-rest).

- Predict class probabilities for test data.
- Compute ROC curves and AUC for each class.
- Plot ROC curves with distinct colors.
```{r}
# Predict class probabilities for test data
rf_probs <- predict(rf_model, newdata = testData, type = "prob")

# Get the true class labels
true_labels <- testData$Crash_Severity

# Define the classes
classes <- levels(true_labels)  # ["Fatal", "Hospitalisation", "Medical treatment", "Minor injury", "Property damage only"]

# Create a list to store ROC objects
roc_list <- list()
auc_values <- numeric(length(classes))

# Compute ROC for each class (One-vs-Rest)
for (i in 1:length(classes)) {
  # Create binary response: 1 for current class, 0 for others
  binary_response <- as.numeric(true_labels == classes[i])
  
  # Get probabilities for the current class
  class_probs <- rf_probs[, classes[i]]
  
  # Compute ROC
  roc_obj <- roc(binary_response, class_probs, levels = c(0, 1), direction = "<")
  roc_list[[classes[i]]] <- roc_obj
  auc_values[i] <- auc(roc_obj)
}

# Plot ROC curves for all classes
plot_colors <- c("#FF6B6B", "#4ECDC4", "#FFD166", "#06D6A0", "#1B9AAA")  # Distinct colors
plot(roc_list[[1]], col = plot_colors[1], main = "Multiclass ROC Curves (One-vs-Rest)")
for (i in 2:length(classes)) {
  plot(roc_list[[i]], col = plot_colors[i], add = TRUE)
}

# Add legend
legend("bottomright", legend = paste(classes, "AUC =", round(auc_values, 3)), 
       col = plot_colors, lwd = 2, cex = 0.8)
```

# Conclusion
- Model Outcome: The Random Forest model achieved low accuracy (22.69%) and high OOB error (70.42%), with poor sensitivity for most classes (e.g., 10.38% for Medical treatment).
- Feature Insights: Involving_Overseas_Licensed_Driver was the most influential predictor (MeanDecreaseAccuracy: 169.58), showing its strong role in crash severity.
- Limitations: Poor performance is due to class imbalance and complex class boundaries, this is further explored in our binary classification model.
  - We suspect the features for non-fatal classes lack sufficient distinctiveness, leading to poor prediction of Hospitalisation, Medical treatment, Minor injury, and Property damage only.
- Value: This analysis provides insights into feature importance and multiclass challenges, and acts as a baseline for the binary classification approach.